---
title: "Mediation analysis with JSmediation"
author: "Cédric Batailler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mediation analysis with JSmediation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Simple mediation

`JSmediation` offers helpers to conduct a joint significant mediation test. 
With this package, you can type a mediation model the way you think it, 
`JSmediation` dealing with fitting underlying models.

In this vignette, we will conduct a joint-significant test for a simple 
mediation scenario.

We will use a data set from Ho, Kteily and Chen (2017). In this experiment, 
authors hypothesized that presenting a text stating that Black-White biracials 
were discriminated would lead Black participants to associate Black-White 
biracials more with their lower status parent group than their higher status 
parent group, according to the rule of hypodescent. In this experiment, the 
authors tested if this effect was mediated by the sense of linked fate between 
discriminated Black-White participants and Black participants. This data set 
is shipped with the `JSmediation` and is called `ho_et_al`.

First, we will load the `JSmediation` package in our environnement so we can use
its functions.

```{r}
library(JSmediation)
```

Now that `JSmediation` is attached, we can load its function, including the 
`ho_et_al` data set.

```{r}
head(ho_et_al)
```

The first thing we can see in this data set is that the `condition` variable is
coded as text something that JSmediation doesn't support. The first step  will 
be to build a new variable which will represent the discrimination
condition with a contrast code. To do so, we will use the `build_contrast` 
function.

```{r}
ho_et_al$condition_c <- build_contrast(ho_et_al$condition,
                                         "Low discrimination",
                                         "High discrimination")

head(ho_et_al)
```

Now that we have a data frame we can use for the analysis, we will use the 
`mdt_simple` function to fit a simple mediation model. Note that any model 
`JSmediation` support is implemented in a `mdt_*` function.

```{r}
my_model <- 
  mdt_simple(ho_et_al,
             IV = condition_c,
             DV = hypodescent,
             M  = linkedfate)
```

We now have an object of class `mediation_model`, we can now whether our data 
support a simple mediation. 

Before looking into the results, we will check for OLS assumptions because 
underlying models used in a joint significant test are actually linear models.

We will extract one model and plot diagnostic plots using the `plot` method for 
object of class `lm`. To do so, we will use the `extract_model` method. To use
this function, we will use the `step` argument which allow us to ask for a 
specific model. Note that `step` allow name of model as well as integer value
to be requested.

```{r}
model_1 <- extract_model(my_model, step = "X -> M")

plot(model_1, ask = FALSE)
```

We will do the same thing for the two other models mdt_simple has fitted.

```{r}
model_2 <- extract_model(my_model, step = 2)

plot(model_2, ask = FALSE)
```

```{r}
model_3 <- extract_model(my_model, step = 3)

plot(model_3, ask = FALSE)
```

It appears that none of the fitted models violate OLS assumptions.

Note that, alternatively, we could have used `extract_models` method to get back
a list of `lm` objects instead of multiple single `lm` objects.

Now that we check for our assumptions, we can interpret our model. To do
so, we just have to call for `my_model`. The structure of objects of class
`mediation_model` is rather complex but there is a `print` method for these
objects. The `print` method is implicitly called one called for the `my_model`
object.

```{r, render="asis"}
my_model
```

In this summary, both a and b paths are significant, we could conclude that 
the indirect effect of discrimination on hypodescent by linked fate is 
significant according to the joint significant test. 

We could choose to compute the Monte Carlo moderation index using `add_index` 
function. This functions adds the indirect effect to the `mediation_model` 
object which is print directly.

```{r}
add_index(my_model)
```

Following this mediation analysis, one could write the following to report the
joint significant mediation test:

> First, we tested examined if  discrimination condition (low vs. high) had an
> effect on hypodescent. This analysis revealed a significant effect, 
> _t_(822) = 2.13, _p_ = .034.
>
> We then examined our hypothesis of interest, namely, that linked fate 
> mediated the effect of discrimination on hypodescent. To do so, we 
> conducted a joint significant test (Yzerbyt et al., 2018). This analysis
> revealed a significant effect of discrimination condition on linked fate,
> _t_(822) = 9.10, _p_ < .001, and a significant effect of linked fate on 
> hypodescent while controlling for discrimination condition, 
> _t_(821) = 5.75, _p_ < .001. The effect of discrimination condition on 
> hypodescent after controlling for linked fate was no longer significant, 
> _t_(821) = 0.33, _p_ = .742.
> Consistently with this analysis, the Monte Carlo confidence interval for the
> indirect effect did not contain 0, CI<sub>95%</sub> [0.0889; 0.208].

# Miscellaneous

Note that `JSmediation` contains some helper function to ease mediation 
analysis.

`display_models` print a summary of each `lm` object fitted during the joint 
significant test.

```{r}
display_models(my_model)
```


`extract_tidy_models` returns a data frame containing models summary 
information.

```{r}
extract_tidy_models(my_model)
```

